Shrinking Characteristics of Precision Matrix Estimators: An Illustration via Regression

### 2019-03-10 04-REGRESS.Rmd
added notations and restructured

# `SCPME` Simulations

## Estimator Comparison

@@

\[ \mathbb{Y} = \mathbb{X}\beta + \mathbb{E} \]

@@

Then for $n$ independent, identically distributed samples:

\[ X_{i} \sim N_{p}\left( 0, \Sigma_{xx} \right) \]

\[ E_{i} \sim N_{r}\left( 0, \Sigma_{y | x} \right) \]

@@

@@

```{r, eval = T, echo = F, fig.cap = "Figure caption here."}

@@

@@

@@

@@

@@

## Shrinking Covariance

We compare the performance of the `shrinkB` and `shrinkBO` estimators when manually shrinking the sample covariance matrix of $X$ and $Y$ (denoted $\Sigma_{xy}$). Each data scenario was replicated a total of 20 times.

@@

\[ \mathbb{Y} = \mathbb{X}\beta + \mathbb{E} \]

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

# Regression Illustration

Talk about regression problem...

## Simulations

@@

\begin{equation}
\mathbb{Y} = \mathbb{X}\beta + \mathbb{E}
(\#eq:datalinregression)\notag
\end{equation}

@@

Then for $n$ independent, identically distributed samples, we have that $X_{i} \sim N_{p}\left( 0, \Sigma_{xx} \right)$ and $E_{i} \sim N_{r}\left( 0, \Sigma_{y | x} \right)$.

@@

```{r scpmesim1, eval = T, echo = F, fig.cap = "Figure caption here."}

@@

```{r scpmesim2, eval = T, echo = F, fig.cap = "Figure caption here."}

@@

```{r scpmesim3, eval = T, echo = F}

@@

```{r scpmesim4, eval = T, echo = F, fig.cap = "Figure caption here."}

@@

```{r scpmesim5, eval = T, echo = F, fig.cap = "Figure caption here."}

@@

```{r scpmesim6, eval = T, echo = F}

@@

### Shrinking Covariance Models

@@

\begin{equation}
\mathbb{Y} = \mathbb{X}\beta + \mathbb{E}
(\#eq:datalinregression2)\notag
\end{equation}

@@

```{r scpmesim7, eval = T, echo = F, fig.cap = "Figure caption here."}

@@

```{r scpmesim8, eval = T, echo = F, fig.cap = "Figure caption here."}

@@

```{r scpmesim9, eval = T, echo = F, fig.cap = "Figure caption here."}

@@

```{r scpmesim10, eval = T, echo = F, fig.cap = "Figure caption here."}

@@

```{r scpmesim11, eval = T, echo = F, fig.cap = "Figure caption here."}

@@

```{r scpmesim12, eval = T, echo = F, fig.cap = "Figure caption here."}

@@

```{r scpmesim13, eval = T, echo = F, fig.cap = "Figure caption here."}

@@

```{r scpmesim14, eval = T, echo = F, fig.cap = "Figure caption here."}

@@

## Discussion

The general optimization problem outlined in @molstad2017shrinking is to estimate $\hat{\Omega}$ such that

\begin{equation}
  \hat{\Omega} = \arg\min_{\Omega \in \mathbb{S}_{+}^{p}}\left\{ tr(S\Omega) - \log\left| \Omega \right| + \lambda\left\| A\Omega B - C \right\|_{1} \right\}
(\#eq:penloglik3)\notag
\end{equation}

where $S$ is the sample covariance of $X \in \mathbb{R}^{p}$ (denominator $n$) and $\left\| \cdot \right\|_{1}$ denotes the $L_{1}$-norm. In addition, $A \in \mathbb{R}^{m \times p}, B \in \mathbb{R}^{p \times q}, \mbox{ and } C \in \mathbb{R}^{m \times q}$ are matrices assumed to be known and specified by the user. This objective function is a penalized log-likelihood for the marginal precision matrix of $X$, which we denote here as $\Omega \equiv \Sigma_{xx}^{-1}$.

\begin{equation}
\hat{\Omega} = \arg\min_{\Omega \in \mathbb{S}_{+}^{p}}\left\{ -l\left( \Omega \right) + P_{\lambda}\left( \Omega \right) \right\}
(\#eq:penloglik4)\notag
\end{equation}

where $P_{\lambda}\left(\Omega \right) = \lambda\left\| A\Omega B - C \right\|_{1}$. Under this particular likelihood, it is assumed that each observation of $X_{i}$ in a sample size of size $n$ follows a $p$-variate normal distribution (ie: $X_{i} \sim N_{p}\left( \mu_{x}, \Omega^{-1} \right)$) -- however, Dennis Cook and others have shown this objective function to work well as a generic loss even when these assumptions are not met.

This particular formulation for the penalty leads to many new, interesting, and novel estimators for $\Omega$. For instance, if in addition to $X$ we also observe $n$ copies of the random variable $Y$, we might set $A = I_{p}, B = \Sigma_{xy}$ where $\Sigma_{xy}$ is the covariance matrix of $X$ and $Y$, and $C = 0$. In which case

\begin{equation}
P_{\lambda}\left(\Omega \right) = \lambda\left\| A\Omega B - C \right\|_{1} = \lambda\left\| \Omega\Sigma_{xy} \right\|_{1} = \lambda\left\| \beta \right\|_{1}
(\#eq:pen3)\notag
\end{equation}

This objective function estimates an $\Omega$ via the marginal log-likelihood of $X$ under the assumption that the forward regression coefficient $\beta$ is sparse (recall that $\beta \equiv \Omega\Sigma_{xy}$). We could also take $B = \left[ \Sigma_{xy}, I_{p} \right]$ so that the identity matrix (dimension $p$) is appended to the covariance matrix of $X$ and $Y$.

\begin{equation}
P_{\lambda}\left(\Omega \right) = \lambda\left\| A\Omega B - C \right\|_{1} = \lambda\left\| \Omega\left[\Sigma_{xy}, I_{p}\right] \right\|_{1} = \lambda\left\| \beta \right\|_{1} + \lambda\left\| \Omega \right\|_{1}
(\#eq:pen4)\notag
\end{equation}

In this case, not only are we assuming that $\beta$ is sparse, but we are also assuming sparsity in $\Omega$. Furthermore, if we let $\mathbb{X} \in \mathbb{R}^{n \times p}$ be a matrix with rows $X_{i} - \bar{X}$ and $\mathbb{Y} \in \mathbb{R}^{n \times r}$ be a matrix with rows $Y_{i} - \bar{Y}$, we might also take $A = \mathbb{X}, B = \Sigma_{xy}, \mbox{ and } C = \mathbb{Y}$ so that

\begin{equation}
\begin{split}
\hat{\Omega} &= \arg\min_{\Omega \in \mathbb{S}_{+}^{p}}\left\{ tr(S\Omega) - \log\left| \Omega \right| + \lambda\left\| \mathbb{X}\Omega \Sigma_{xy} - \mathbb{Y} \right\|_{1} \right\} \\
  &= \arg\min_{\Omega \in \mathbb{S}_{+}^{p}}\left\{ tr(S\Omega) - \log\left| \Omega \right| + \lambda\left\| \mathbb{Y} - \mathbb{X}\beta \right\|_{1} \right\}
(\#eq:penloglik5)\notag
\end{split}
\end{equation}

This estimator encourages the predicted values $\hat{\mathbb{Y}} = \mathbb{X}\hat{\beta} = \mathbb{X}\hat{\Omega}\Sigma_{xy}$ to be exactly equal to the observed responses $\mathbb{Y}$. We will name this estimator the *lasso SCPME estimator*. Note that in practice - for any of these estimators - we do not know the true covariance matrix $\Sigma_{xy}$ but we could provide the sample estimate $\hat{\Sigma}_{xy} = \mathbb{X}^{T}\mathbb{Y}/n$ in its place.

All of these estimators fall into the SCPME (shrinking characteristics of precision matrix estimators) algorithmic framework outlined in the paper. The augmented ADMM algorithm (details [here](https://mgallow.github.io/SCPME/articles/Details.html#augmented-admm-algorithm)) is the following:

Set $k = 0$ and repeat steps 1-6 until convergence.

1. Compute $G^{k} = \rho A^{T}\left( A\Omega^{k} B - Z^{k} - C + \Lambda^{k}/\rho \right)B^{T}$

2. Decompose $S + \left( G^{k} + (G^{k})^{T} \right)/2 - \rho\tau\Omega^{k} = VQV^{T}$ (via the spectral decomposition).

3. Set $\Omega^{k + 1} = V\left( -Q + (Q^{2} + 4\rho\tau I_{p})^{1/2} \right)V^{T}/(2\rho\tau)$

4. Set $Z^{k + 1} = \mbox{soft}\left( A\Omega^{k + 1}B - C + \rho^{-1}\Lambda_{k}, \rho^{-1}\lambda \right)$

5. Set $\Lambda^{k + 1} = \rho\left( A\Omega^{k + 1} B - Z^{k + 1} - C \right)$

6. Replace $k$ with $k + 1$.

### 2019-03-10 06-ADMMsigma-Package.Rmd
added notations and restructured

`ADMMsigma` is an R package that estimates a penalized precision matrix, often denoted $\Omega$, via the alternating direction method of multipliers (ADMM) algorithm. Though not the fastest estimation method, the ADMM algorithm is easily adaptable and allows for rapid experimentation by the user -- the primary goal of this package.

@@

* `ADMMsigma()` computes the estimated precision matrix (ridge, lasso, and elastic-net type regularization optional)

@@

* `plot.ADMMsigma()` produces a heat map or line graph for cross validation errors

* `plot.RIDGEsigma()` produces a heat map or line graph for cross validation errors

@@

<br>\vspace{0.5cm}
```{r, eval = FALSE}
# The easiest way to install is from CRAN
install.packages("ADMMsigma")

@@

This package is hosted on Github at [github.com/MGallow/ADMMsigma](https://github.com/MGallow/ADMMsigma).

@@

Here, the data matrix is $100 \times 5$ and generated from a multivariate normal distribution with mean zero and tapered oracle covariance matrix $S$. A tapered covariance matrix has an inverse - or precision matrix - that is tri-diagonal. That is, the precision matrix is sparse and the shrinkage penalty used here should prove useful.

@@

We can also explicitly assume sparsity in our estimate - rather than let the package decide - by restricting the class of penalties to the lasso. We do this by simply setting `alpha = 1` in our function:

@@

`ADMMsigma` also has the capability to provide plots for the cross validation errors. This allows the user to analyze and select the appropriate tuning parameters. In the heatmap plot below, the more bright (white) areas of the heat map correspond to a better tuning parameter selection. In the line graph, each line corresponds to a different $\alpha$ tuning parameter.

@@

`ADMMsigma` has a number of more advanced options such as cross validation criteria, regularization path, and parallel processing that are explained in detail on the project website at [mattxgalloway.com/ADMMsigma](http://mattxgalloway.com/ADMMsigma/).

>>>>

`ADMMsigma` is an R package that estimates a penalized precision matrix, often denoted $\Omega$, via the alternating direction method of multipliers (ADMM) algorithm. Though not the fastest estimation method, the ADMM algorithm is easily adaptable and allows for rapid experimentation by the user, which is the primary goal of this package.

@@

* `ADMMsigma()` computes the estimated precision matrix via the ADMM algorithm (ridge, lasso, and elastic-net regularization optional)

@@

* `plot.ADMMsigma()` produces a heat map or optional line graph for cross validation errors

* `plot.RIDGEsigma()` produces a heat map or optional line graph for cross validation errors

@@

```{r, eval = FALSE}
# The easiest way to install is from CRAN
install.packages("ADMMsigma")

@@

This package is hosted on Github at [github.com/MGallow/ADMMsigma](https://github.com/MGallow/ADMMsigma). The project website is located at [mattxgalloway.com/ADMMsigma](http://mattxgalloway.com/ADMMsigma/).

@@

Here, we will use a $100 \times 5$ data matrix that is generated from a multivariate normal distribution with mean zero and tapered oracle covariance matrix $S$. A tapered covariance matrix has an inverse - or precision matrix - that is tri-diagonal, which is useful for illustration purposes because the object is sparse (many zero entries) and the shrinkage penalty should prove useful.

@@

We can also explicitly assume sparsity in our estimate rather than let the package decide by restricting the class of penalties to the lasso. We do this by simply setting `alpha = 1` in our function:

@@

`ADMMsigma` also has the capability to provide plots for the cross validation errors. This allows the user to analyze and manually select the appropriate tuning parameters. In the heatmap plot below, the bright, white areas of the heat map correspond to a better tuning parameter selection (low validation error). In the line graph, each line corresponds to a different $\alpha$ tuning parameter.

@@

`ADMMsigma` has a number of more advanced options such as cross validation criteria, regularization path, and parallel CV that are explained in more detail on the project website at [mattxgalloway.com/ADMMsigma](http://mattxgalloway.com/ADMMsigma/).

### 2019-03-10 07-SCPME-Package.Rmd
added notations and restructured

where $\lambda > 0$ is a tuning parameter, $A, B, \mbox{ and } C$ are known matrices specified by the user, and we define $\left\|A \right\|_{1} = \sum_{i, j} \left| A_{ij} \right|$.

This form of penalty leads to many new, interesting, and novel estimators for the precision matrix $\Omega$. In short, users can construct matrices $A, B, \mbox{ and } C$ so that emphasis is place on the sum, absolute value of a *characteristic* of $\Omega$. We will explore a few of these in the tutorial section.

@@

* `plot.shrink()` produces a heat map or line graph for cross validation errors

@@

<br>\vspace{0.5cm}
```{r, eval = FALSE}
# The easiest way to install is from CRAN
install.packages("SCPME")

@@

This package is hosted on Github at [github.com/MGallow/SCPME](https://github.com/MGallow/SCPME).

@@

The primary function is the `SCPME` package is simply `shrink()`. The input values $X$ is an $n \times p$ data matrix so that there are $n$ rows each representing an observation and $p$ columns each representing a unique variable and $Y$ is an $n \times r$ response matrix where $r$ is the dimension of the response vector. By default, `SCPME` will estimate $\Omega$ using a lasso penalty ($A = I_{p}, B = I_{p}, \mbox{ and } C = 0$) and choose the optimal $\lambda$ tuning parameter that minimizes the mean squared prediction error for the regression of the variable $Y$ on $X$. If $Y$ is not provided (optional), then tuning parameter selection will be based on the validation likelihood. Note that $\Omega$ (perhaps better denoted here as $\Omega_{x}$) will only have meaningful shrinkage unless the data vector $X \in \mathbb{R^{p}}$ is multi-dimensional - that is, $p > 1$.

Here, in this example, the data matrix is $100 \times 5$ and the response is generated according to the following model:

@@

where $E_{i} \sim N\left( 0, 1 \right)$ and $X_{i}$ is generated from a multivariate normal distribution with mean zero and tapered oracle covariance matrix $S$. A tapered covariance matrix has an inverse - or precision matrix - that is tri-diagonal. It is this oracle precision matrix that is our primary interest in estimating. In addition, $\beta$ is randomly generated and sparse. The data will be generated using the `data_gen()` function contained in the package.

@@

# generate data from a sparse oracle precision matrix.
# we can use the built-in `data_gen` function

@@

```{r, message = FALSE, echo = TRUE}

@@

Notice here that the estimated precision matrix is *not* sparse. This is due to the fact that our cross validation criteria is the mean-squared prediction. We can estimate a new precision matrix using the validation likelihood as criteria with the following command:

@@

Recall that all of the estimators so far have used a lasso penalty that penalizes the sum of the absolute value of all the entries in $\Omega$ ($A = I_{p}, B = I_{p}, \mbox{ and } C = 0$). In effect, this penalty embeds an assumption in our estimate that the true $\Omega$ is sparse. The flexibility of the penalty described in @molstad2017shrinking allows us to make other assumptions as well. For instance, in the penalty we could set $A = I_{p}, B = \Sigma_{xy}$ where $\Sigma_{xy}$ is the covariance matrix of $X$ and $Y$, and $C = 0$. In which case

@@

Note that we specified the maximum `lam` value in the previous function to expand the tuning parameter grid. Conveniently, with these settings, the augmented ADMM algorithm also solves for the estimated $\beta$ coefficient matrix simultaneously:

@@

Another possible penalty is to set $B = \left[ \Sigma_{xy}, I_{p} \right]$ so that the identity matrix (dimension $p$) is appended to the covariance matrix of $X$ and $Y$. That is, the penalty $P$, is constructed as follows:

@@

In this case, not only are we assuming that $\beta$ is sparse, but we are also assuming sparsity in $\Omega$.

@@

`SCPME` also has the capability to provide plots (heatmap and line graph) for the cross validation errors. In the heatmap plot below, the more bright (white) areas of the heat map correspond to a better tuning parameter selection.

@@

`SCPME` has a number of more advanced options such as various convergence criteria and parallel processing that are explained in detail on the project website at [mattxgalloway.com/SCPME](http://mattxgalloway.com/SCPME/).

>>>>

where $\lambda > 0$ is a tuning parameter, $A, B, \mbox{ and } C$ are known, user-specified matrices, and we define $\left\|A \right\|_{1} = \sum_{i, j} \left| A_{ij} \right|$.

This form of penalty leads to many new, interesting, and novel estimators for the precision matrix $\Omega$. Users can construct matrices $A, B, \mbox{ and } C$ so that emphasis is placed on the sum, absolute value of a *characteristic* of $\Omega$. We will explore a few of these estimators in the tutorial section.

@@

* `plot.shrink()` produces a heat map or optional line graph for the cross validation errors

@@

```{r, eval = FALSE}
# The easiest way to install is from CRAN
install.packages("SCPME")

@@

This package is hosted on Github at [github.com/MGallow/SCPME](https://github.com/MGallow/SCPME). The project website is located at [mattxgalloway.com/SCPME](http://mattxgalloway.com/SCPME/).

@@

The primary function in the `SCPME` package is `shrink()`. The input values $X$ is an $n \times p$ data matrix so that there are $n$ rows each representing an observation and $p$ columns each representing a unique variable and $Y$ is an $n \times r$ response matrix where $r$ is the dimension of the response vector. By default, `SCPME` will estimate $\Omega$ using a lasso penalty ($A = I_{p}, B = I_{p}, \mbox{ and } C = 0$) and choose the optimal $\lambda$ tuning parameter that minimizes the mean squared prediction error for the regression of the variable $Y$ on $X$ (here $I_{p}$ denotes a $p$-dimension identity matrix). If $Y$ is not provided, then tuning parameter selection will be based on the validation likelihood. Note that $\Omega$ (perhaps better denoted here as $\Omega_{x}$) will only have meaningful shrinkage unless the data vector $X \in \mathbb{R^{p}}$ is multi-dimensional ($p > 1$).

In this example, the data matrix is $100 \times 5$ and the response is generated according to the following model:

@@

where $E_{i} \sim N\left( 0, 1 \right)$ and $X_{i}$ is generated from a multivariate normal distribution with mean zero and tapered oracle covariance matrix $S$. A tapered covariance matrix has the property that its inverse - the precision matrix - is tri-diagonal. Estimating this oracle precision matrix well and efficiently will be our primary interest. In addition, $\beta$ is randomly generated and sparse. The data will be generated using the `data_gen()` function contained in the package.

@@

@@

@@

Notice here that the estimated precision matrix is *not* sparse. This is due to the fact that our cross validation criteria is the mean-squared prediction error. We can estimate a new precision matrix using the validation likelihood as the cross validation criteria with the following command:

@@

All of the estimators so far have used a lasso penalty that penalizes the sum of the absolute value of all the entries in $\Omega$ ($A = I_{p}, B = I_{p}, \mbox{ and } C = 0$). In effect, this penalty embeds an assumption in our estimate that the true population precision matrix, $\Omega$, is sparse. The flexibility of the penalty described in @molstad2017shrinking allows us to make other assumptions as well. For instance, in the penalty we could set $A = I_{p}, B = \Sigma_{xy}$ where $\Sigma_{xy}$ is the covariance matrix of $X$ and $Y$, and $C = 0$. In which case our penalty function

@@

Note that we specified the maximum `lam` value in the previous function to expand the tuning parameter grid. With these settings, the augmented ADMM algorithm also solves for the estimated $\beta$ coefficient matrix simultaneously:

@@

Another possible penalty is to set $B = \left[ \Sigma_{xy}, I_{p} \right]$ so that the identity matrix is appended to the covariance matrix of $X$ and $Y$. That is, the penalty $P$, is constructed as

@@

In this case, we are equally penalizing the sum, absolute values of entries in $\beta$ *and* $\Omega$ which embeds an assumption that $\beta$ and $\Omega$ are both sparse.

@@

`SCPME` also has the capability to provide plots (heatmaps and line graphs) for the cross validation errors. In the heatmap plot below, the more bright, white areas correspond to a better tuning parameter selection (lower cross validation error).

@@

`SCPME` has a number of more advanced options including alternative convergence criteria and parallel CV that are explained in detail on the project website at [mattxgalloway.com/SCPME](http://mattxgalloway.com/SCPME/).
